Clearance Project - 3 - TKstats

### 2025-04-03 data-raw/ClearanceProject-3-TKstats.Rmd
Added hepatic blood flow sensitivity analysis as a possible test for restrictive/nonrestrictive clearance assumption.

```{r library-loading}
today <- format(Sys.Date(), "%d%b%Y")
devtools::load_all()
library(dplyr)
library(tidyr)
library(stringr)
library(ggplot2)
library(ctxR)
library(ggdist)
library(patchwork)
library(httk)
```

@@

Next I want to categorize each chemical based on putative hepatic extraction ratio.
This is calculated as:
$$
E_H = \frac{f_{up}\cdot CL_{int}}{Q_{hep} + f_{up}\cdot CL_{int}}
$$
```{r extraction_ratio}
rat_er <- rpr_param_df %>%
  group_by(Chemical) %>%
  reframe(fupClint = (Funbound.plasma * Clint),
          Qhep = (Qcardiacc*Qliverf)) %>%
  mutate(extraction_ratio = fupClint/(Qhep + fupClint),
         clearance_type_by_er = case_when(
           extraction_ratio >= 0.7 ~ "Nonrestrictive",
           extraction_ratio < 0.3 ~ "Restrictive",
           between(extraction_ratio, 0.3, 0.7) ~ "Ambiguous"
         )
  )

@@

>>>>

```{r library-loading}
today <- format(Sys.Date(), "%d%b%Y")
devtools::load_all()
library(dplyr)
library(tidyr)
library(stringr)
library(ggplot2)
library(ctxR)
library(ggdist)
library(patchwork)
library(httk)
load_dawson2021()

@@

Next I want to categorize each chemical based on putative hepatic extraction ratio.
This is calculated as:

$$
E_H = \frac{f_{up}\cdot CL_{int}}{Q_{hep} + f_{up}\cdot CL_{int}}
$$
```{r extraction_ratio}
rat_er <- rpr_param_df %>%
  group_by(Chemical) %>%
  reframe(fupClint = (Funbound.plasma * Clint),
          Qhep = (Qcardiacc*Qliverf)) %>%
  mutate(extraction_ratio = fupClint/(Qhep + fupClint),
         clearance_type_by_er = case_when(
           extraction_ratio >= 0.7 ~ "Nonrestrictive",
           extraction_ratio < 0.3 ~ "Restrictive",
           between(extraction_ratio, 0.3, 0.7) ~ "Ambiguous"
         )
  )

@@

I want to do some sensitivity analysis for hepatic blood flow.
The logic here is that, with nonrestrictive clearance, the clearance is
flow-dependent and therefore would be sensitive to increases in hepatic
blood flow. Restrictive compounds should not.

For rats, the default hepatic blood flow (in mL/min/kg^(3/4)) is 33.38.
I will increase this by 1% (i.e. 0.3338) and look at the sensitivity of
steady-state concentration and total clearance.

```{r hbf_sensitivity}
# Find 1% increment
liver_tissue_new <- subset(tissue.data,
                           Tissue == "liver" &
                             Species == "Rat")

new_hepblfl_increment <- liver_tissue_new[
  liver_tissue_new$variable == "Flow (mL/min/kg^(3/4))",
  "value"
] * (1/100) # Get the increment amount (this is delta r)
# Calculate numerator for normalized sensitivity coefficient, 1%
drr <- new_hepblfl_increment/liver_tissue_new[
  liver_tissue_new$variable == "Flow (mL/min/kg^(3/4))",
  "value"
]

# Decrease "rest" of body blood flow
tissue.data[
  tissue.data$Tissue == "rest" &
    tissue.data$Species == "Rat" &
    tissue.data$variable == "Flow (mL/min/kg^(3/4))",
  "value"
  ] <- tissue.data[
  tissue.data$Tissue == "rest" &
    tissue.data$Species == "Rat" &
    tissue.data$variable == "Flow (mL/min/kg^(3/4))",
  "value"
  ] - new_hepblfl_increment
# Increase hepatic blood flow
tissue.data[
  tissue.data$Tissue == "liver" &
    tissue.data$Species == "Rat" &
    tissue.data$variable == "Flow (mL/min/kg^(3/4))",
  "value"
  ] <- 
  tissue.data[
  tissue.data$Tissue == "liver" &
    tissue.data$Species == "Rat" &
    tissue.data$variable == "Flow (mL/min/kg^(3/4))",
  "value"
  ] + new_hepblfl_increment
```

```{r}
# Next we need to re-parameterize the chemicals...
# I only need to do this under the assumption of restrictive clearance.
# The test is that under this assumption these chemicals should NOT be sensitive
# to the changes in hepatic blood flow.

rpr_sparams_rest <- tryCatch(
  expr = {
    sapply(
      target_chems_rat,
      \(x) {
        httk::parameterize_3comp2(
          dtxsid = x,
          species = "Rat",
          restrictive.clearance = TRUE,
          default.to.human = FALSE,
          force.human.clint.fup = FALSE,
          suppress.messages = TRUE
        )
      },
      simplify = FALSE,
      USE.NAMES = TRUE
    )
  }, error = function(msg) {
    message("Not possible to parameterize_3comp2 for all dtxsid ",
            "try to use default.to.human = TRUE.")
  }
)
# For these same chemicals, I also want to test sensitivity
# when parameterized as human for fup and Clint
rph_sparams_rest <- tryCatch(
  expr = {
    sapply(
      target_chems_rat,
      \(x) {
        httk::parameterize_3comp2(
          dtxsid = x,
          species = "Rat",
          restrictive.clearance = TRUE,
          default.to.human = TRUE,
          force.human.clint.fup = TRUE,
          suppress.messages = TRUE
        )
      },
      simplify = FALSE,
      USE.NAMES = TRUE
    )
  }, error = function(msg) {
    message("Not possible to parameterize_3comp2 for all dtxsid ",
            "try to use default.to.human = TRUE.")
  }
)

# Copied from RC_comparison_bkgd.R
css_httk_batch <- function(param_list,
                           restrictive = TRUE) {
  stopifnot(is.logical(restrictive))

  css_list_r <- vapply(
    param_list,
    \(x) {
      calc_analytic_css_3comp2(parameters = x,
                               restrictive.clearance = restrictive)
    },
    FUN.VALUE = numeric(1L)
  )
  css_list_names <- names(param_list)

  css_df <- data.frame(Chemical = css_list_names,
                       Css_httk_r.1p = css_list_r)
  return(css_df)
}

rpr_css_sensitivity <- css_httk_batch(rpr_sparams_rest)
rph_css_sensitivity <- css_httk_batch(rph_sparams_rest)

rpr_cltot_sensitivity <- data.frame(
      Chemical = names(rpr_sparams_rest),
      CLtot_restrictive.1p = vapply(
        rpr_sparams_rest,
        \(x) {
          httk::calc_total_clearance(
            parameters = x,
            species = "Rat",
            model = "3compartment2",
            suppress.messages = TRUE,
            restrictive.clearance = TRUE
          )
        },
        FUN.VALUE = double(1)
      ),
      halflife_restrictive.1p = vapply(
        rpr_sparams_rest,
        \(x) {
          httk::calc_half_life(
            parameters = x,
            species = "Rat",
            model = "3compartment2",
            suppress.messages = TRUE,
            restrictive.clearance = TRUE
          )
        },
        FUN.VALUE = double(1)
      )
)

rph_cltot_sensitivity <- data.frame(
      Chemical = names(rph_sparams_rest),
      CLtot_restrictive.1p = vapply(
        rph_sparams_rest,
        \(x) {
          httk::calc_total_clearance(
            parameters = x,
            species = "Rat",
            model = "3compartment2",
            suppress.messages = TRUE,
            restrictive.clearance = TRUE
          )
        },
        FUN.VALUE = double(1)
      ),
      halflife_restrictive.1p = vapply(
        rph_sparams_rest,
        \(x) {
          httk::calc_half_life(
            parameters = x,
            species = "Rat",
            model = "3compartment2",
            suppress.messages = TRUE,
            restrictive.clearance = TRUE
          )
        },
        FUN.VALUE = double(1)
      )
)

rpr_sensitivity <- merge(rpr_cltot_sensitivity,
                         rpr_css_sensitivity)

rph_sensitivity <- merge(rph_cltot_sensitivity,
                         rph_css_sensitivity)

rpr_sensitivity <- rpr_sensitivity %>%
  merge(rpr_httk_clearance_df) %>%
  mutate(
    nsc_css = ((Css_httk_r.1p - 
                  Css_httk_r)/Css_httk_r)/drr,
    nsc_cltot = ((CLtot_restrictive.1p -
                    CLtot_restrictive)/CLtot_restrictive)/drr,
    nsc_halflife = ((halflife_restrictive.1p -
                       halflife_restrictive)/halflife_restrictive)/drr
  ) %>%
  select(Chemical, Species, starts_with("nsc_"))

rph_sensitivity <- rph_sensitivity %>%
  merge(rpr_httk_clearance_df) %>%
  mutate(
    nsc_css = ((Css_httk_r.1p - 
                  Css_httk_r)/Css_httk_r)/drr,
    nsc_cltot = ((CLtot_restrictive.1p -
                    CLtot_restrictive)/CLtot_restrictive)/drr,
    nsc_halflife = ((halflife_restrictive.1p -
                       halflife_restrictive)/halflife_restrictive)/drr
  ) %>%
  select(Chemical, Species, starts_with("nsc_"))

rpr_sensitivity %>%
  mutate(across(starts_with("nsc_"),
                \(x) {
                  factor(
                    case_when(
                      abs(x) < 0.2 ~ 'Low',
                      between(abs(x), 0.2, 0.5) ~ 'Medium',
                      abs(x) > 0.5 ~ 'High'
                    ), levels = c('High', 'Medium', 'Low')
                  )
                })) %>% 
  count(nsc_css, nsc_cltot, nsc_halflife)

addmargins(clearance_contingency_tbl)

rph_sense_tables <- rph_sensitivity %>%
  mutate(across(starts_with("nsc_"),
                \(x) {
                  factor(
                    case_when(
                      abs(x) < 0.2 ~ 'Low',
                      between(abs(x), 0.2, 0.5) ~ 'Medium',
                      abs(x) > 0.5 ~ 'High'
                    ), levels = c('High', 'Medium', 'Low')
                  )
                })) %>%
  rename_with(\(x) {paste0(x, "_h")}, .cols = starts_with("nsc_"))

rpr_sense_tables <- rpr_sensitivity %>%
  mutate(across(starts_with("nsc_"),
                \(x) {
                  factor(
                    case_when(
                      abs(x) < 0.2 ~ 'Low',
                      between(abs(x), 0.2, 0.5) ~ 'Medium',
                      abs(x) > 0.5 ~ 'High'
                    ), levels = c('High', 'Medium', 'Low')
                  )
                })) %>% 
  left_join(combo_pretable) %>% 
  arrange(nsc_css, nsc_cltot, nsc_halflife) %>%
  relocate(rat_cl, human_cl, .after = Species) %>%
  mutate(
    cl = paste0("Rat:", rat_cl, "/Human:", human_cl),
  .keep = "unused", .after = Species) %>%
  left_join(rph_sense_tables) %>%
  left_join(fup_rh)

with(rpr_sense_tables,
     table(cl, nsc_halflife))

rpr_sense_tables %>%
  mutate(Rat_ambi = str_detect(cl, "Rat:Ambi")) %>%
  filter(!is.na(nsc_css)) %>%
  ggplot(aes(x = Clint_human, 
             y = Clint_rat, 
             color=cl)) +
  geom_point(alpha = 0.7, aes(shape = Rat_ambi)) +
  scale_shape_manual(values = c(16,21)) +
  scale_color_manual(
    values = c(
      "#d6604d", "red2", "#b2182b",
      "#fdb863", "#e08214", "#b35806",
      "#b2abd2", "#8073ac", "#542788")
  ) +
  labs(x = "Human", y = "Rat",
       title = "CL<sub>int</sub>",
       color = "Clearance Determination\nby Hepatic Extraction",
       shape = "Rat Clearance\ndetermined Ambiguous?") +
  facet_grid(rows = vars(nsc_halflife),
             scales = "free",
             space = "free_x") +
  geom_abline(slope = 1, color = "grey20") +
  theme_bw() +
  theme(plot.title = ggtext::element_markdown(hjust = 0.5))

ggsave(filename = paste0(Sys.getenv("FIG_DIR"),
                         "RestrictiveClearance_Analysis/",
                         "Rat_Human_Clint_Clearance_by_extractionRatio.png"),
       height = 7, width = 8,
       bg = "white")

Unfortunately, it seems that the sensitivity for Css, CLtot, and halflife
when parameterized with human fup and CLint values is quite high,
meaning they don't provide as much information as other metrics.
